---
title: "group_by"
output:
  github_document:
    toc: true
author: Guojing Wu
data: 2018-10-04
---

```{r setup, include=FALSE}
library(tidyverse)
library(rnoaa)
library(ggridges)
library(patchwork)
knitr::opts_chunk$set( # default setting
  echo = TRUE, 
  fig.width = 6, 
  fig.asp = 0.6, 
  out.width = "90%")
devtools::install_github("thomasp85/patchwork")
theme_set(theme_bw() + theme(legend.position = "bottom"))
```

data import

```{r data import, cache=TRUE}
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, date, month, everything())
```

## start grouping

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  ungroup()
```

doing some summarize

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(number = n())

weather_df %>% 
  count(name, month)
```

Compute multiple summaries within each group, the difference between summarize and mutate is that when you do group_by, summarize is gonna collapse all the observations, but mutate is gonna show all the observations and simply add another column.

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(number = n(),
            num_na = sum(is.na(tmax)), # count missing data
            mean_tmax = mean(tmax, na.rm = T), 
            sd_tmin = sd(tmin, na.rm = T))

weather_df %>%
  group_by(name, month) %>%
  mutate(mean = mean(tmax, na.rm = T))
```

## now let's plot normal ggplot

```{r}
weather_df %>%
  group_by(name, month) %>%
  summarize(number = n(),
            num_na = sum(is.na(tmax)), # count missing data
            mean_tmax = mean(tmax, na.rm = T), 
            sd_tmin = sd(tmin, na.rm = T)) %>% 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point()

```

ggplot using different dataset

```{r}
month_df = 
  weather_df %>%
  group_by(name, month) %>%
  summarize(number = n(),
            num_na = sum(is.na(tmax)), # count missing data
            mean_tmax = mean(tmax, na.rm = T), 
            sd_tmin = sd(tmin, na.rm = T))

ggplot(weather_df, aes(x = date, y = tmax, color = name)) +
  geom_point(alpha = 0.5) +
  geom_line(data = month_df, aes(x = month, y = mean_tmax, color = name))
```

Sometimes tidy date is not helpful

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarise(mean_tmax = mean(tmax, na.rm = T)) %>% 
  spread(key = name, value = mean_tmax)
```

doing some centralization

```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(mean_tmax = mean(tmax, na.rm = T), 
         centered_tmax = tmax - mean_tmax) %>% 
  ggplot(aes(x = date, y = centered_tmax, color = name)) +
  geom_point()
```

## window funtions

use ranks to see the coldest day in each month

```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(tmax_rank = min_rank(tmax)) %>% 
  filter(tmax_rank < 2)
```

reverse the result

```{r}
weather_df %>%
  group_by(name, month) %>%
  mutate(tmax_rank = min_rank(desc(tmax))) %>% 
  filter(tmax_rank < 2)
```

using lag

```{r}
weather_df %>%
  group_by(name) %>%
  mutate(lag = lag(tmax))
```

## how heat is this?

```{r}
weather_df %>%
  group_by(name) %>%
  mutate(one_day_change = tmax - lag(tmax)) %>% 
  summarize(sd_one_day_change = sd(one_day_change, na.rm = T))
```

